prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-27 inst/MyArticle/paper.Rmd
Add changes to Modelling, Results and Conclusion

However, if there is no observation in a group, the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean, using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample (see Supplement section Estimating $A_{LOD}$). Typically there are many such cases, and hence we take the median.

@@

Where, $n_g$ is the number of groups and $n$ is the number of observations.
If variance can not be estimated for a protein, because there are to few observations in other groups, we use the median pooled variance of all other proteins in the dataset.

@@

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom $df$. @Smyth2004linear discuss how, to use the $\sigma$ and $df$ of all models to estimate a prior $\sigma$ and prior $df$, and posterior $\tilde \sigma$. These can be used to moderate the $t$-statistics by:

@@

To summarize peptide level models to protein models, we apply the method suggested by @Suomi2017bEnhanced that use the median scaled $p$-value of the peptide models and the cumulative distribution function of the Beta distribution (CDF) to determine a regulation probability of the protein.

To obtain the $\tilde{p}$ of a protein, we first rescaled the peptide $p$-values by taking the sign of the fold-change $\hat \beta$ into account, i.e.:

@@

Afterwards, the median scaled $p$-value $\tilde{p}_s$ is determined and using the transformation below, transformed back onto the original scale:

@@

Because we use the median, with the i-th order statistic $i = \frac{n}{2} + 0.5$, we parametrize the CDF of the Beta distribution with $\gamma = i = \frac{n}{2} + 0.5$ and $\delta = n - i + 1 = n - (\frac{n}{2} + 0.5) + 1 = \frac{n}{2} + 0.5 = \gamma$. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

@@

The following code example illustrates how we compute differences among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `merge_contrasts_results` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:figure2). Panel A shows the distribution of the $p$-values, Panel B shows the volcano plot for each comparison, and Panel C shows a Bland-Altman plot which shows the difference between groups as a function of the rank of the protein abundance.

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method to convert the `LFQData` class into a `r BiocStyle::Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code, where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")`s functionality (see Supplementary Figure S[X]).

To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the prolfqua package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric,@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula and which comparisons to compute from the sample annotation provided in tabular form, similarly to SAINTexpress and `r BiocStyle::Biocpkg("MSstats")`. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, `r BiocStyle::Githubpkg("fgcz/prolfqua")` and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and helps scientists to meet requirements from funding agencies, journals, and academic institutions to publish data according to the FAIR (Findable, Accessible, Interoperable, and Reusable) [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with results produced by `r BiocStyle::Biocpkg("MSstats")`, `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes all modelling methods we evaluated.

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimated differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

Figure \@ref(fig:figure3) Panel B shows how various estimates obtained from the models, i.e., the difference between groups, $t$-statistics, and scaled $p$-values allow identifying true positives (TP) given a false positive rate (FPR) of $10%$, by displaying the partial area under the ROC. Ordering proteins by the $t$-statistic or $p$-value leads to a higher $pAUC_{10}$ than when ordering by the estimated difference among groups. Suppose an accurate estimate of the difference among groups is essential. In that case, the linear models fitted to protein intensities, calculated using Tukey's median polish, perform better than those directly modeling peptide intensities, e.g., `ropeca` or `prot_mixed_effect_moderated` (see Table \@ref(tab:describeBenchmarkedModels) Abundance column). We speculate that outliers at the peptide level do not affect the protein estimates when using Tukey's median polish method, a non-parametric method to infer protein abundances. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

@@

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:figure3) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, indicating that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

@@

The IonStar dataset (see Methods) captures only the variance from the chromatography, electro-spray, and mass spectrometric measurements since only technical replicates are available for each dilution. Therefore essential sources of variation typically present in other experiments, such as the biochemical and biological, are not measured. Furthermore, this dataset with a parallel group design does not allow for benchmarking models with interactions. However, other benchmark datasets have similar constraints [@edwards2015cptac, @o2018proteome]. Furthermore, we could not analyze the benchmark dataset [@wessels2012comprehensive] with a complete factorial design with an up-to-date version of MaxQuant. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret our findings. Specifically, the observed variances will be higher in datasets with biological replicates, and the power will be lower for the same number of samples. Furthermore, the proportion of missing observations in real-life datasets might be higher or distributed differently in groups.

We can conclude that if we want to sort the proteins according to the likelihood of being differentially regulated to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values improves the inference (see Figure \@ref(fig:figure3) panel B).
There is no such improvement for the mixed effect model, because the degrees of freedom are erroneously estimated. Furthermore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by @Smyth2004linear, consistently improves the ranking of proteins compared with the unmoderated estimates (not shown). However, since also for this method, a correct degree of freedom estimate is required, it does not work for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:figure3) Panel A left). Furthermore, by using the Beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:figure3) Panel A center). However, the properties of Beta-based probabilities are not well understood; for instance, the p-values are not uniformly distributed under the null hypothesis. Furthermore, the $FDR$ estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased and overestimate the false discovery proportion (see Figure \@ref(fig:figure3) Panel C). Therefore, we can not recommend this method if an unbiased estimate of $FDR$ is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:figure3) Panel C).

@@

Since, the API of `r BiocStyle::Biocpkg("proDA")` and the produced outputs are similar to those of prolfqua, we are integrating the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

We focused our benchmark on comparing the statistical modeling methods using three different scores while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing a differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r BiocStyle::Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast], used to identify and quantify proteins, significantly contributes to the entire pipeline's performance, altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the supplement we show the benchmarking results for `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Githubpkg("fgcz/prolfqua")` when the IonStar dataset is prepossessed with the FragPipe v15 [@yu2020fast] quantification pipeline. Although the number of proteins identified with two peptides is slightly lower than with MaxQuant (3929 vs. 4046), the pAUC_{10} is significantly higher ($81%$ vs. $72%$).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is a feature-rich, object-oriented R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)). Furthermore, the contrast specification is explicit and consistent for all models and allows for testing interactions. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. Furthermore, the developed framework can integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables the implementation of applications where the user can select an alternative normalization method, protein abundance estimation method, or differential expression analysis algorithm. Furthermore, this R package can analyze other types of quantitative proteomics data,e.g., labeling-based TMT-, PRM- DIA-data.

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the $FDR$ is an unbiased estimate of the $FDP$. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice: i) estimate protein abundances from peptide abundances using a robust or nonparametric regression method; ii) use linear models because they show good performance in all categories; iii) if the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead; iv) if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates; v) If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference; and last but not least vi) do not impute missing observation but statistically model missingness to estimate parameters, i.e., group differences. Finally, the differential expression analysis results obtained with prolfqua are comparable to those of other differential expression analysis tools.

>>>>

However, if there is no observation in a group the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample (see Supplement section Estimating $A_{LOD}$). Typically there are many such cases, and hence we take the median.

@@

Where $n_g$ is the number of groups and $n$ is the number of observations.
If variance can not be estimated for a protein, because there are too few observations in other groups, we use the median pooled variance of all other proteins in the dataset.

@@

From the linear and the mixed effect models, we can obtain the residual standard deviation $\sigma$, and degrees of freedom $df$. @Smyth2004linear discuss how, to use the $\sigma$ and $df$ of all models to estimate the corresponding priors and posterior $\tilde \sigma$. These can be used to moderate the $t$-statistics by:

@@

To summarize peptide level models to protein models, we apply the method suggested by @Suomi2017bEnhanced that uses the median scaled $p$-value of the peptide models and the cumulative distribution function of the Beta distribution (CDF) to determine a regulation probability of the protein.

To obtain the $\tilde{p}$ of a protein we first rescaled the peptide $p$-values by taking the sign of the fold-change $\hat \beta$ into account, i.e.:

@@

Afterwards, the median scaled $p$-value $\tilde{p}_s$ is determined and, using the transformation below, transformed back onto the original scale:

@@

Because we use the median with the i-th order statistic $i = \frac{n}{2} + 0.5$, we parametrize the CDF of the Beta distribution with $\gamma = i = \frac{n}{2} + 0.5$ and $\delta = n - i + 1 = n - (\frac{n}{2} + 0.5) + 1 = \frac{n}{2} + 0.5 = \gamma$. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

@@

The following code example illustrates how we compute differences among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `merge_contrasts_results` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:figure2). Panel A shows the distribution of the $p$-values, Panel B shows the volcano plot for each comparison, and Panel C shows a Bland-Altman plot reporting the difference between groups as a function of the rank of the protein abundance.

@@

Using the data frame of tidy data ensures interoperability with other proteomics-related packages that manage their data with tidy-tables, e.g., `r BiocStyle::CRANpkg("protti")` [@quast2022protti]. To simplify the integration of `r BiocStyle::Githubpkg("fgcz/prolfqua")` with Bioconductor based workflows there is a method that converts the `LFQData` class into a `r BiocStyle::Biocpkg("SummarizedExperiment")`. The use of R6 classes, which encapsulate the configuration and the data, allows for writing very concise code where functions can have few arguments. Auto-completion support for R6 classes in the Rstudio editor makes it easy for novices to explore `r BiocStyle::Githubpkg("fgcz/prolfqua")'s functionality (see Supplementary Figure S[X]). 
To ease the usage barriers of the R package to users not familiar with statistics and R programming, we developed an application based on the prolfqua package into our data management platform [B-Fabric](https://fgcz-bfabric.uzh.ch/) [@bfabric,@bfabricjib2022]. The B-Fabric system runs a computing infrastructure controlled by a local resource management system that supports cloud-bursting [@VMMAD]. This integration enables users to select the input data and basic settings in a graphical user interface (GUI).

In this GUI application, we determine the model formula and which comparisons to compute from the sample annotation provided in tabular form, similarly to SAINTexpress and `r BiocStyle::Biocpkg("MSstats")`. However, this application can analyze only parallel group designs with and without repeated measurements or factorial designs without interactions. The user receives a report including input files, the R markdown file, and R scripts necessary to replicate the analysis using their in-house R installation. In this way, `r BiocStyle::Githubpkg("fgcz/prolfqua")` and [B-Fabric](https://fgcz-bfabric.uzh.ch/) and helps scientists to meet requirements from funding agencies, journals, and academic institutions while publishing their data according to the FAIR (Findable, Accessible, Interoperable, and Reusable) [@FAIR] data principles. We are working on creating a shiny stand-alone application with the described functionality and making it available soon.

@@

The Benchmark functionality of `r BiocStyle::Githubpkg("fgcz/prolfqua")` includes receiver operator curves (ROC) and computes partial areas under those curves ($pAUC$) and other scores, e.g., the false discovery proportion $FDP$. We use those scores, i.e. the $pAUC$ at $10%$ $FDR$ and the $FDP$, to examine how well the methods implemented in `r BiocStyle::Githubpkg("fgcz/prolfqua")` model quantitative mass spectrometric high throughput data and compare them with the results produced by `r BiocStyle::Biocpkg("MSstats")`, `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Biocpkg("msqrob2")`. Table \@ref(tab:describeBenchmarkedModels) summarizes all the modelling methods we evaluated.

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimates differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

Figure \@ref(fig:figure3) Panel B shows how various estimates obtained from the models, i.e., the difference between groups, $t$-statistics and scaled $p$-values, allow identifying true positives (TP) given a false positive rate (FPR) of $10%$ by displaying the partial area under the ROC. Ordering proteins by the $t$-statistic or $p$-value leads to a higher $pAUC_{10}$ than when ordering by the estimated difference among groups. Suppose an accurate estimate of the difference among groups is essential. In that case, the linear models fitted to protein intensities, calculated using Tukey's median polish, perform better than those directly modeling peptide intensities, e.g., `ropeca` or `prot_mixed_effect_moderated` (see Table \@ref(tab:describeBenchmarkedModels) Abundance column). We speculate that outliers at the peptide level do not affect the protein estimates when using Tukey's median polish method. This hypothesis is examined in the benchmark, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`.

@@

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:figure3) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, which indicates that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

@@

The IonStar dataset (see Methods) captures only the variance from the chromatography, electro-spray, and mass spectrometric measurements since only technical replicates are available for each dilution. Therefore essential sources of variation typically present in other experiments, such as the biochemical and biological ones, are not measured. Furthermore, this dataset with a parallel group design does not allow for benchmarking models with interactions. However, other benchmark datasets have similar constraints [@edwards2015cptac, @o2018proteome]. Furthermore, we could not analyze the benchmark dataset [@wessels2012comprehensive] with a complete factorial design with an up-to-date version of MaxQuant. Thus, while we can extrapolate some of the results obtained to more realistic datasets, we need to be careful not to over-interpret our findings. Specifically, the observed variances will be higher in datasets with biological replicates, and the power will be lower for the same number of samples. Furthermore, the proportion of missing observations in real-life datasets might be higher or distributed differently in groups.

We can conclude that if we want to sort the proteins according to the likelihood of being differentially regulated to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values improves the inference (see Figure \@ref(fig:figure3) panel B).
There is no such improvement for the mixed effect model, because the degrees of freedom are erroneously estimated. Furthermore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by @Smyth2004linear, consistently improves the ranking of proteins compared with the unmoderated estimates (not shown). However, since also for this method a correct degree of freedom estimate is required, it does not work for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:figure3) Panel A left). Furthermore, by using the Beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:figure3) Panel A center). However, the properties of Beta-based probabilities are not well understood; for instance, the p-values are not uniformly distributed under the null hypothesis. Furthermore, the $FDR$ estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased and overestimate the FDP (see Figure \@ref(fig:figure3) Panel C). Therefore, we can not recommend this method if an unbiased estimate of $FDR$ is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:figure3) Panel C).

@@

Since the API of `r BiocStyle::Biocpkg("proDA")` and the produced outputs are similar to those of prolfqua, we are integrating the `r BiocStyle::Biocpkg("proDA")` package as an additional modeling option into `r BiocStyle::Githubpkg("fgcz/prolfqua")` (see Figure \@ref(fig:ContrastUML)).

We focused our benchmark on comparing the statistical modeling methods using three different scores while we fixed the pre-processing steps. However, some of these steps are of utmost significance when performing a differential expression analysis [@frohlich2022benchmarking]. One of them is the normalization of the intensities within the samples to remove systematic differences [@pursiheimo2015optimization]. The method used to infer proteins from peptide identifications [@nesvizhskii2005interpretation] and protein abundances from peptide abundances is an additional important factor [@Grossmann2010]. For instance, the original `r BiocStyle::Biocpkg("proDA")` publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using MaxLFQ intensities reported by MaxQuant, the $pAUC_{10}$ is lower ($pAUC_{10}(\textrm{t-statistics}) = 66%$) compared with results obtained when protein abundances are estimated from peptides abundances using Tukey's median polish ($pAUC_{10}(\textrm{t-statistics}) = 72%$). Last but not least, the software [@Cox2008MaxQuant; @yu2020fast] used to identify and quantify proteins significantly contributes to the entire pipeline's performance altering the number of identified proteins and the sensitivity and specificity of the differential expression analysis. In the Supplement we show the benchmarking results for `r BiocStyle::Biocpkg("proDA")` and `r BiocStyle::Githubpkg("fgcz/prolfqua")` when the IonStar dataset is prepossessed with the FragPipe v15 [@yu2020fast] quantification pipeline. Although the number of proteins identified with two peptides is slightly lower than with MaxQuant (3929 vs. 4046), the pAUC_{10} is significantly higher ($81%$ vs. $72%$).

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is a feature-rich, object-oriented R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)). Furthermore, the contrast specification is explicit and consistent for all models and allows for testing interactions. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. Furthermore, the developed framework can integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables the implementation of applications where the user can select an alternative normalization method, protein abundance estimation methods, or differential expression analysis algorithms. Furthermore, this R package can analyze other types of quantitative proteomics data,e.g., labeling-based TMT-, PRM- DIA-data.

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the $FDR$ is an unbiased estimate of the $FDP$. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computational experiments, we can provide the following advice: i) estimate protein abundances from peptide abundances using a robust or nonparametric regression method; ii) use linear models because they show good performance in all categories; iii) if the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead; iv) if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates; v) If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference; and last but not least vi) do not impute missing observation but statistically model missingness to estimate parameters, i.e., group differences. Finally, the differential expression analysis results obtained with prolfqua are comparable to those of other differential expression analysis tools.
